---
title: "Quantifying biodiversity effects across times and places"
output: html_notebook
---

# Bioversity-functioning partitions

```{r include = FALSE}
# load relevant libraries
library(here)
library(dplyr)

# load the functions for the partition
source(here("scripts/Isbell_2018_partition.R"))

# load the test data
t.dat <- readRDS(here("test_data/Isbell_test_data.rds"))
```

## Relative yield

What is overyielding? Overyielding occurs when the production of a species in mixture exceeds the expected production of that species based on its monoculture production. This indicates that, for some reason, mixtures of species are greater than the sum of their parts (i.e. their monocultures). Therefore, we often want to quantify overyielding behaviour to determine if there are benefits to mixing multiple species together.

This brings us to *relative yield*. Relative yield does not measure overyielding directly. Rather, we use relative yield as a tool to quantify overyielding. Specifically, we quantify relative yield as the mixture yield of species i divided by the monoculture yield of species i. Thus, for the observed relative yield, we quantify it as:

* RYobs,i = Yobs,i/Mi 

Where **RYobs,i** is the relative yield of species i, **Yobs,i** is the observed yield in mixture of species i and **Mi** is the monoculture yield of species i.

If we want to know whether there is overyielding, we have to compare the observed relative yield to something. If our mixture is sown with equal proportions of two species, then the *expected yield* of species i (Yexp,i) would be half the monoculture value of species i. Therefore, the expected relative yield of species (RYexp,i) would be 0.5. This is because the expected yield of species i is half its monoculture. This can be seen mathematically as:

* RYexp,i = 0.5 = Yexp,i/Mi
* Yexp,i = 0.5*Mi

In this case, if the observed relative yield (RYobs,i) exceeds the expected relative yield (RYexp,i) then species i in mixture has a higher yield than expected based on the monoculture yield (Mi) and the sown proportions (i.e. intial proportion). In experimental data, we simply use the sown proportion as our RYe. In natural systems, we will have to be a bit more clever. Probably, we will either have to assume a distribution of possible initial proportions (i.e. drawing from the Dirichlet distribution) or use multiple time-points.

The relative yield calculations are the bread and butter of the original Loreau and Hector (2001) partition and Isbell et al.'s (2018) extension. Specifically, we measure the net biodiversity effect NBE:

* sum( delta RYi * Mi)

Where delta RYi is the difference between RYobs,i and RYexp,i for species i. This NBE is then partitioned into two different terms in Loreau and Hector's (2001) partition, the complementarity effect (CE) and the selection effect (SE):

* CE = N mean( delta RYi ) mean( Mi ) 
* SE = N cov( delta RYi, Mi )

The selection effect occurs when species with higher than average monoculture yields also have higher relative yields than expected. Complementarity effects occur when species yields are on average higher than expected based on their monoculture yields.


## Isbell et al.'s (2018) biodiversity effects

*Local complementarity effects can become selection effects at larger scales*

Local complementarity effects can become selection effects at larger scales. How does this occur? Take the example below as a test case. Basically, you have the situation where two species have equal monoculture yields in two different places. But, the monoculture yields are much greater in one place than the other. In addition, at the place where the monoculture yields are more productive, they both overyield. At the place where the monoculture yields are lower, they underyield:

```{r}
# load the test data
df <- t.dat[[1]]
print(df)
```

If we then calculate local complementarity effects, we get strong local complementarity at the first place but negative complementarity effects at the second place. 

```{r}
# calculate the complementarity effect

# calculate change in relative yield
df$RY_obs <- df$Y/df$M

# input expected relative yield
df$RY_exp <- rep(c(0.5, 0.5), 2)

# calculate change in relative yield
df$dRY <- df$RY_obs - df$RY_exp

# calculate the complementarity effect for each sample
CE_loc <- 
  df %>%
  group_by(sample) %>%
  summarise(CE = n() * mean(dRY) * mean(M) )
 
print(CE_loc)
 
```

If we sum these to get overall local complementarity effects, we get a positive effect i.e. a positive local complementarity effect across all places.

```{r}
print(paste("Local CE across places: ", sum(CE_loc$CE)))
```
At local scales, there cannot be any selection effects because the monoculture yields are equal which precludes any covariance:

```{r}
# calculate covariance within each site: always zero
raw_cov(df[1:2,]$dRY, df[1:2,]$M)
raw_cov(df[3:4,]$dRY, df[3:4,]$M)

```
However, if we calculate the total selection effect across times and places, we see a positive total selection effect. This is because species overyield most where they are most productive in monoculture. Specifically, there is a positive correlation between the change in relative yield and their monoculture which leads to a total selection effect across places.

```{r}
# calculate the total selection effect
N <- n_unique(df$place) * n_unique(df$time) * n_unique(df$species)
N * raw_cov(df$dRY, df$M)

```

*Local selection effects can become complementarity effects at larger scales*

We have local selection effects when there is covariance between overyielding (i.e. change in relative yield) and monoculture yields. To see how these local selection effects can become complementarity effects at larger scales, consider the following example. In this example, we have two species at two environments. One species is more productive in monoculture in both environments. But, in the productive environment, the high monoculture yield species overyields most. In the low productivity environment, the low monoculture yield species overyields most. The data are:

```{r}
# load the test data
df <- t.dat[[2]]
print(df)
```

What happens when we calculate local complementarity effects? As we can see below, there is a weak complementarity effect at both places. Therefore, when we add them up, they cancel each other out.

```{r}
# calculate the complementarity effect

# calculate change in relative yield
df$RY_obs <- df$Y/df$M

# input expected relative yield
df$RY_exp <- rep(c(0.5, 0.5), 2)

# calculate change in relative yield
df$dRY <- df$RY_obs - df$RY_exp

# calculate the complementarity effect for each sample
CE_loc <- 
  df %>%
  group_by(sample) %>%
  summarise(CE = n() * mean(dRY) * mean(M) )
 
print(CE_loc)
 
```

Similarly, if we calculate total selection effects, they also sum to zero. Why does this occur? This occurs because, at the large scale, the fact that the low productivity species overyields most in the other place deouples the monoculture, change in relative yield correlation.

```{r}
# calculate total selection effects
print(df)
plot(df$M, df$dRY)

N <- n_unique(df$place) * n_unique(df$time) * n_unique(df$species)
N * raw_cov(df$dRY, df$M)

```

But, if we calculate local selection effects, there is a strong positive local selection effect in the productive environment. Therefore, that leads to a total complementarity effect which can be interpreted as a type of spatial niche partitioning.

```{r}
# calculate local selection effects
n_unique(df$species) * raw_cov(df[1:2, ]$dRY, df[1:2,]$M)
n_unique(df$species) * raw_cov(df[3:4, ]$dRY, df[3:4,]$M)

```

*Space-time partition: Overview*

To extend this partition, they split the total selection effect into two components. 

Total insurance effect: PTN * cov(dPobs,ijk, Mijk)

This quantifies the extent to which the most productive speies in monoculture tend to dominate mixtures. Theory predicts that it should be positive when:

* When the best competitor for a single limiting resource outcompetes all others (Tilman et al. 1997)

* In spatially or temporally fluctuating environments where species tend to dominate in places (Loreau et al. 2003) and times (Yachi and Loreau 1999) where they are most productive in monoculture

Non-random overyielding

The non-random overyielding term is quite confusing because it's not exactly clear what the delta RYobs,ijk term measures. The explanation in the text says that it is the extent to which the most productive species in monoculture tend to overyield most in mixtures. They then say that overyielding is quantified relative to observed rather than initial relative abundance... The formula is:

* PTN cov( delta RYobs,ijk, Mijk)

My initial explanation is that if there is a high RYobs,ijk but the species is dominating in the mixture (i.e. high Pobs,ijk), then this is not non-random overyielding because we it is not overyielding necessarily more than we expect based on its proportion?

Total complementarity effect

This term quantifies the extent to which species overyield on average. Therefore, it quantities the extent to which niche partitioning and facilitation outweight negative effects of chemical warfare etc.


*Six types of biodiversity effect*

**Case 1: Average selection effect**

This is the case of a homogeneous and temporally constant environment. In this case, there are two species and one has higher monoculture yield (species 1). Moreover, it is the best competitor so it will eventually fully dominate the mixtures and species 2 will be out-competed. Under this scenario, the mixture yield matches the yield of the dominant monoculture.

In this case, the net biodiversity effect is 100 and it is completely due to the average selection effect.

**Case 2: Temporal insurance effect**

This case occurs when species exhibit asynchronous responses to environmental fluctuations and are able to competitively dominate at times when the conditions are most favourable to them (i.e. high monoculture yield), (Yachi and Loreau 1999).

Like case 1, mixtures become monocultures of the most productive species but the identity of this most productive species changes over time. This could only occur if species change their dominance structure in response to changes in the environment. Therefore, species must compete over relevant timescales.

* Timescales are another important implication of theory...

This effect assumes that species can be lost at one time and then recolonise somehow. However, if it goes globally extent, then we expect a permanent loss in functioning.

**Case 3: Spatial insurance effect**

This case occurs when species exhibit asynchronous responses to spatial heterogeneity and are able to competitively dominate at times when the conditions are most favourable to them (i.e. high monoculture yield), (Yachi and Loreau 1999).

Like the temporal insurance effects, mixtures of species become dominated by the most productive species in monoculture but the identity of that species differs from year to year.

**Case 4: Spatio-temporal insurance effect**

This is similar to temporal and spatial insurance effects. However, it describes cases where we cannot attribute covariation between monoculture yields and mixture relative biomass to either time or space.

**Case 5: Complementarity effect**

Even if you have a temporally and spatially homogeneous environment, different species can still partition the niche by, for example, using different combinations of limiting nutrients, growing at different times of the year etc.

Complementarity effects can also arise via facilitation whereby the presence of one species actually improves the growth of other species.

Under this scenarios, ecosystem functioning depends on both species at all times and all places.

**Case 6: Complementarity with non-random overyielding**

This quantifies the extent to which the most productive species in monoculture overyields most in mixture. This is more of a residual effect and doesn't have a clear theoretical interpretation yet.

They expect that this effect will probably be negative although it's not very clear why...


## Does the partition behave as expected given inputs from a simple metacommunity model?

I've implemented a modified version of the model presented in Thompson et al. (2020, Ecology Letters). The modification was to implement proper Lotka-Volterra competition rather than the Beverton-Holt competition. This means that species have K-values for a given patch as well as r-values for a given patch. I use the same function to generate the K-values as the r-values which means that species that have high grow rates in a patch in some place at some time also have high K-values. This is probably a realistic assumption but it can be modified if necessary.

This modification makes it easier to create interpretable competition coefficient matrices unlike the Thompson et al. (2020) model which has to use a scaling factor to increase equilibrium abundances.

I then use this model to set-up simulations of BEF-experiments at metacommunity scales. Specifically, I've used **five species** in **five patches**. The species have different environmental optima and narrow niche widths. Specifially, I have created the idealised situation where species niches are distributed as shown below. Notes that the environment in this model can only vary between 0 and 1.

```{r echo = FALSE}
# calculate total selection effects
print( data.frame(species = 1:5,
                  env_optimum = c(0.1, 0.3, 0.5, 0.7, 0.9), 
                  env_niche_width = c(0.2)) 
       )

```


All species are seeded at equal abundance into each patch in the "mixture" treatment. In addition, I create all possible monocultures of the five species at the metacommunity scale. Therefore, in the species 1 monoculture, species 1 is seeded into all five patches and no new species can enter the system.

To test whether the partition performs as expected, I created different scenarios that I implement and then I quantify biodiversity effects as per Isbell et al. (2018).

**Environmental variation**

I consider three types of environmental variation:

* 1. Pure spatial variation: Each patch has a different environmental value but the patches do not vary in time

* 2. Pure temporal variation: All patches have the same environmental value but this varies perfectly synchronously in time in all patches

* 3. Combined spatial and temporal variation: Patches vary in both space and time in their environmental value

**Competition**

I consider two types of competition:

* 1. Neutral competition: All competition coefficients are one which means that, in the absence of dispersal and possible mass effects, there should be zero or very little local complementarity among species.

* 2. Stabilising competition: Competition coefficients vary between zero and one which means that local complementarity can arise. Competition is assumed to be symmetric and intraspecific competition is assumed to be one.

I use combinations of these effects on a common landscape template and with equal connectivity to explore whether Isbell et al.'s (2018) partition makes intuitive sense given the assumptions implemented in the model.


```{r include = FALSE}

# Load the functions and packages
library(ggplot2)
library(tidyr)
library(dplyr)
library(here)
source(here("scripts/mcomsimr_simulate_MC2_function.R"))

# set the seed
set.seed(5489485)


## Set-up fixed inputs
species <- 3
patches <- 3
ts <- 100


## Landscape parameters

# generate a random landscape
l.1 <- 
  data.frame(x = c(50, 25, 25),
             y = c(50, 25, 75))
plot(l.1)

# generate a random dispersal matrix
d.1 <- dispersal_matrix(landscape = l.1, torus = TRUE, kernel_exp = 0.1, plot = FALSE)


## Species traits

# generate species environmental optima
t.1 <- 
  data.frame(species = 1:species,
             optima = seq(0.1, 0.9, length.out = species),
             env_niche_breadth = 0.2,
             max_r = 0.5,
             K_max = 150)
head(t.1)


## Competition matrices

# generate species interaction matrix randomly

# matrix 1: Intraspecific > interspecific competition
si.1 <- matrix(runif(n = species*species, min = 0.1, max = 1), 
                  nrow = species, ncol = species)
si.1[lower.tri(si.1)] = t(si.1)[lower.tri(si.1)]
diag(si.1) <- 1
head(si.1)

# matrix 2: Neutral competition
si.2 <- species_int_mat(species = species, intra = 1, min_inter = 1, 
                        max_inter = 1, comp_scaler = 1, plot = FALSE)
head(si.2)


## Environmental matrices

# generate a random environmental matrix
e.1 <- env_generate(landscape = l.1, env1Scale = 900, timesteps = (ts), plot = FALSE )

# only spatial variation
e.2 <- expand.grid(1:patches, 1:(ts))
e.2$env1 <- rep(seq(0.1, 0.9, length.out = species), ts ) 
names(e.2) <- c("patch", "time", "env1")
e.2 <- e.2[,c("env1", "patch", "time")] 
head(e.2)


# autocorrelated temporal variation only: Temporal insurance effect
x <- env_generate(landscape = landscape_generate(patches = 2, plot = FALSE), 
                  env1Scale = 5, timesteps = (ts), plot = FALSE )
e.3 <- expand.grid(1:patches, 1:ts)
e.3$env1 <- rep(x[x$patch == 1, ][1:ts, ]$env1, each = patches)
names(e.3) <- c("patch", "time", "env1")
e.3 <- e.3[,c("env1", "patch", "time")] 

# only random temporal variation i.e. strong average selection effect
e.4 <- expand.grid(1:patches, 1:ts)
e.4$env1 <- rep(sample(x = seq(0.1, 0.9, 0.1), size = ts, replace = TRUE), 
                each = patches ) 
names(e.4) <- c("patch", "time", "env1")
e.4 <- e.3[,c("env1", "patch", "time")] 


# set-up a function to explore these biodiversity effects under different assumptions
Isbell_explorer <- function(env, comp, n_runs, disp_rate, t_sel) {

x_out <- vector("list", length = n_runs)
BEF_out <- vector("list", length = n_runs)
for (i in 1:n_runs) {
  x <- 
    sim_metacomm_BEF(patches = patches, species = species, dispersal = disp_rate, 
                     timesteps = ts, start_abun = 150,
                     extirp_prob = 0.001,
                     landscape = l.1, disp_mat = d.1, env.df = env, 
                     env_traits.df = t.1, int_mat = comp)
  
  x_out[[i]] <- x
  
  BEF_out[[i]] <- Isbell_2018_cleaner(mix = x$mixture, mono = x$monoculture, t_sel  = t_sel)
  
}

# plot the results

p1 <- 
  x_out[[sample(1:length(x_out), 1)]]$mixture %>%
  mutate(species = as.character(species)) %>%
  ggplot(data = .,
         mapping = aes(x = time, y = N, colour = species)) +
  geom_line(size = 1) +
  scale_colour_viridis_d() +
  facet_wrap(~patch, scales = "free") +
  theme_classic()
  
p2 <- 
  bind_rows(BEF_out, .id = "run") %>%
  pivot_longer(cols = c("NBE", "TC", "TS", "AS", "TI", "SI", "ST", "LC"),
               names_to = "BEF_effect",
               values_to = "value") %>%
  ggplot(data = ., 
         mapping = aes(x = BEF_effect, y = value)) +
  geom_jitter(alpha = 0.1, width = 0.1) +
  geom_boxplot(width = 0.1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic()
  
plot(p1)
plot(p2)

}



```


*Average selection effect*

First, we simulate the effect of the most productive species in monoculture dominating the mixture.

```{r echo = FALSE, fig.height = 2, fig.width = 6}

# temporal variation only: e.4
# neutral competition: si.2
# no dispersal: disp_rate = 0.05

Isbell_explorer(env = e.4, comp = si.2, n_runs = 10, disp_rate = 0, t_sel = seq(10, ts, 20))
  
```


*Pure spatial variation*

Next, we simulate the case of pure spatial variation, neutral competition and no dispersal. In this case, we expect that all biodiversity effects would be due to spatial insurance. There is no local complementarity due to species interactions (neutral competition) and no mass effects because dispersal is zero.


```{r echo = FALSE}

# spatial variation only: e.2
# neutral competition: si.2
# no dispersal: disp_rate = 0

Isbell_explorer(env = e.2, comp = si.2, n_runs = 10, disp_rate = 0, t_sel = (ts-5):ts )
  
```


Next, we simulate the same situation but with some dispersal. This means that there can be some mass effects which leads to some limited local complementarity.

```{r echo = FALSE}

# spatial variation only: e.2
# neutral competition: si.2
# some dispersal: disp_rate = 0.05

Isbell_explorer(env = e.2, comp = si.2, n_runs = 10, disp_rate = 0.05, t_sel = (ts-5):ts)
  
```


Finally, we simulated pure spatial variation and local complementarity by using a competition coefficient matrix that, for interspecific competition, is less than one.

```{r echo = FALSE}

# spatial variation only: e.2
# stabilising competition: si.1
# dispersal: disp_rate = 0.05

Isbell_explorer(env = e.2, comp = si.1, n_runs = 10, disp_rate = 0.05, t_sel = (ts-5):ts)
  
```


*Pure temporal variation*

Now, we look at cases of pure temporal variation. This means that all patches have the same environmental conditions in space but vary in time. We start with the case of neutral competition and some dispersal.

```{r echo = FALSE}

# temporal variation only: e.3
# neutral competition: si.2
# no dispersal: disp_rate = 0.05

Isbell_explorer(env = e.3, comp = si.2, n_runs = 10, disp_rate = 0, t_sel = seq(10, ts, 20))
  
```

Next, we examine the same situation but with stabilising competition. Therefore, the local complementarity effects should get stronger.


```{r echo = FALSE}

# temporal variation only: e.3
# stabilising competition: si.1
# no dispersal: disp_rate = 0.1

Isbell_explorer(env = e.3, comp = si.1, n_runs = 10, disp_rate = 0.1, t_sel = seq(10, ts, 20))
  
```

```{r echo = FALSE}

# temporal variation only: e.3
# neutral competition: si.1
# no dispersal: disp_rate = 0.1

Isbell_explorer(env = e.3, comp = si.2, n_runs = 10, disp_rate = 0.3, t_sel = seq(10, ts, 20))
  
```


*Mixed environmental variation*

Now we look at cases where the environment varies in both space and time and there is neutral competition. Here, we expect there to be both spatial and temporal insurance effects but limited local complementarity because competition is neutral. We use some dispersal here.

```{r echo = FALSE}

# spatial and temporal variation: e.1
# neutral competition: si.2
# no dispersal: disp_rate = 0.05

Isbell_explorer(env = e.1, comp = si.2, n_runs = 10, disp_rate = 0.05, t_sel = seq(10, ts, 20))
  
```


In all cases, there are no temporal insurance effects. This seems to happen because the time-scales that we are looking at are so short. Let's try to look at time-points that are more spread out so that there is more opportunity for communities to adjust.




